Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (ACC with TMM and SVA normlaisation)

### 2016-09-22 code/Rmd/CMC_ACC_StarAligned_TMM_SVA.Rmd
Modified ACC pipeline with SVA to account for change in covariates

---
title: "Covariate analysis of CMC RNASeq data reprocessed with Star-HTSeq pipeline (ACC with TMM normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knitr)
library(knit2synapse)
library(synapseClient)

@@

knit2synapse::knitToFolderEntity(file = "./CMC_ACC_StarAligned_TMM.Rmd",
                                 parentId ="syn5569976",
                                 entityName = 'ACC - STAR Reprocessed - with TMM and SVA')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5569976';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of star aligned CMC reprocessed RNASeq data with TMM normalisation (ACC)';

@@

## Metadata with specific information for RNASeq processing
metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Ethnicity", "Age of Death", "PMI (in hours)", 
                          "pH", "Dx", "CDR", "CauseOfDeath", "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
                          "ACC_RNA_isolation: RNA Isolation Batch", "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?",
                          "ACC_RNA_report: Library Batch")

@@

METADATA = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Ethnicity, Age_of_Death, PMI__in_hours_, pH, Dx, CDR, CauseOfDeath,
                ACC_RNA_isolation__Exclude_, ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RNA_Isolation_Batch,
                ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_, ACC_RNA_report__Library_Batch) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = ACC_RNA_isolation__Exclude_, 
                SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
                IsolationBatch = ACC_RNA_isolation__RNA_Isolation_Batch,
                RIN = ACC_RNA_isolation__RIN, 
                ReportExclude = ACC_RNA_report__Exclude_, 
                LibraryBatch = ACC_RNA_report__Library_Batch) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  droplevels() %>%
  dplyr::mutate(LibraryBatch = gsub('[^[:alnum:]]','_',LibraryBatch),
                LibraryBatch = as.factor(LibraryBatch),
                IsolationBatch = as.factor(IsolationBatch))
```

### Data preprocessing
```{r preprocess.data, echo=TRUE}
# Remove samples marked as exclude and samples with no Gender, Ethnicity, PMI, and RIN
METADATA <- METADATA %>%
  filter(SampleID %in% colnames(COUNT)) %>%
  filter(IsolationExclude == 0, ReportExclude == 0) %>%
  filter(!is.na(PMI)) %>%
  filter(!is.na(Gender)) %>%
  filter(!is.na(Ethnicity)) %>%
  filter(!is.na(RIN)) %>%
  filter(!(SampleID %in% c('MSSM_RNA_ACC_250', 'MSSM_RNA_ACC_170'))) %>%
  as.data.frame()
```

@@

### Normalisation (Custom design)
Since many covariates are correlated, re-normalising COUNTS with a custom design matrix. Here "Dx" is chosen as the primary variable of interest
```{r custom.norm}
# Primary variable of interest
postAdjustCovars = c('Dx', 'RIN', 'RIN2', 'Age_of_Death', 'PMI', 'LibraryBatch')

@@

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
Estimate surrogate variables using SVA package
```{r sva}
MODEL0 = DM1$design[,-grep('Dx',colnames(DM1$design))] # Get null model by removing variable of interest
MODEL1 = DM1$design

>>>>

---
title: "Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (ACC with TMM and SVA normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knitr)
library(knit2synapse)
library(synapseClient)

@@

knit2synapse::knitToFolderEntity(file = "./CMC_ACC_StarAligned_TMM_SVA.Rmd",
                                 parentId ="syn5569976",
                                 entityName = 'ACC - STAR Reprocessed - with TMM and SVA')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5569976';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of star aligned CMC reprocessed RNASeq data with TMM and SVA normalisation (ACC)';

@@

## Metadata with specific information for RNASeq processing
metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", 
                          "Dx", "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?", "ACC_RNA_report: Library Batch",
                          'DNA_genotyping_report: Genotyping Sample_ID')

@@

METADATA = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, ACC_RNA_isolation__Exclude_,
                ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_,
                ACC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = ACC_RNA_isolation__Exclude_, 
                SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
                RIN = ACC_RNA_isolation__RIN, 
                ReportExclude = ACC_RNA_report__Exclude_, 
                LibraryBatch = ACC_RNA_report__Library_Batch,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  droplevels() %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch))

# Replace AFF and BP samples as other
levels(METADATA$Dx) = c("Other", "Other", "Control", "SCZ")
```

### Data preprocessing
```{r preprocess.data}
# Remove samples marked as exclude and samples with no Gender, Ethnicity, PMI, and RIN
writeLines('Following counts are missing any metadata')
writeLines(paste(setdiff(colnames(COUNT), levels(METADATA$SampleID)), collapse = ', '))
METADATA <- METADATA %>% filter(SampleID %in% colnames(COUNT))

writeLines('Following samples are marked exclude')
writeLines(paste(METADATA$SampleID[METADATA$IsolationExclude == 1 | METADATA$ReportExclude == 1], collapse = ', '))
METADATA <- METADATA  %>% filter(IsolationExclude == 0, ReportExclude == 0)

writeLines('Following samples are missing PMI information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$PMI)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(PMI))

writeLines('Following samples are missing gender information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$Gender)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(Gender))

writeLines('Following samples are missing RIN information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$RIN)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(RIN)) 
```

@@

### Normlaisation (Custom design)
To be consistent with ACC Nature Neuroscience article we performed a custom adjustment with Dx, RIN, RIN2, Age of Death, PMI, Library Batch, Institution, Gender
```{r custom.design}
# Add primary variable to the list (in case not chosen by iterative method)
postAdjustCovars = c('Dx', c(FactorCovariates, ContCovariates)) %>% unique()

@@

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Estimate surrogate variables using SVA package
```{r sva}
MODEL0 = DM1$design[,-grep('Dx',colnames(DM1$design))] # Get null model by removing variable of interest
MODEL1 = DM1$design
